---
title: "Statistical Inference Course Project"
author: "Jonas Wiorek"
date: "Thursday, April 23, 2015"
output: html_document
---

- Title (give an appropriate title) and Author Name
- Overview: In a few (2-3) sentences explain what is going to be reported on.
- Simulations: Include English explanations of the simulations you ran, with the accompanying R code. Your explanations should make clear what the R code accomplishes.
- Sample Mean versus Theoretical Mean: Include figures with titles. In the figures, highlight the means you are comparing. Include text that explains the figures and what is shown on them, and provides appropriate numbers.
- Sample Variance versus Theoretical Variance: Include figures (output from R) with titles. Highlight the variances you are comparing. Include text that explains your understanding of the differences of the variances.
- Distribution: Via figures and text, explain how one can tell the distribution is approximately normal.

```{r}
lambda <- 0.2
n <- 40
nosim <- 1000
rexp(n,lambda)
mu <- 1/lambda
sigma <-1/lambda 
mean(rexp(n,lambda))
var(rexp(n,lambda))

hist(apply(matrix(rexp(nosim*n, lambda),nosim),1,var))


# Sample Mean
# The population mean mu is 1/lambda for a population with exponential distribution with rate lambda 
lambda
mu
hist(apply(matrix(rexp(nosim*n, lambda),nosim),1,mean))
# The expexcted value of the sample mean, xHat, is the population mean mu
xHat <- mean(apply(matrix(rexp(nosim*n, lambda),nosim),1,mean))
xHat
mu
# The variance of the sample mean is sigma^2/n
sdMean <- sd(apply(matrix(rexp(nosim*n, lambda),nosim),1,mean))
sdMean^2
sigma^2/n
# The logical estimate of the variance of the sample mean is S^2/n
S <- mean(apply(matrix(rexp(nosim*n, lambda),nosim),1,sd))
S^2/n



```

Assume we have a population of exponential distribution with a  mean and standard deviation equal to lambda.
n samples of this exponential distribution has mean asymptotically equal to lambda.
The mean of n samples of this exponetial distribution has a standard deviation asymptotically equal lambda/sqrt(n).

# Sample Mean
# The population mean mu is is 1/lambda for a population with exponential distribution with lambda 
# The expexcted value of the sample mean, xHat, is the population mean mu
# The variance of the sample mean is sigma^2/n
# The logical estimate of the variance of the sample mean is s^2/n ???

# Sample Variance
# The sample variance, S^2, estimates the population variance sigma^2
# The sample variance expected value is the population variance
# The distribution of the sample variance is centered around sigma^2

```{r, echo=FALSE}

# Sample Variance
# The sample variance, S^2, estimates the population variance sigma^2
# The sample variance expected value is the population variance
# The distribution of the sample variance is centered around sigma^2
hist(apply(matrix(rexp(nosim*n, lambda),nosim),1,var))
mean(apply(matrix(rexp(nosim*n, lambda),nosim),1,var))
S^2
sigma^2

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
